---
title: "Project 7: Difference-in-Differences and Synthetic Control"
authors: Jose Aveldanes, Monica De La Cruz, Lawrence Liu
output: html_document
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth)

# set seed
set.seed(1)

# load data
medicaid_expansion <- read_csv('medicaid_expansion.csv')
```

# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?
- Which states were home to most uninsured Americans prior to 2014? How about in 2021? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

```{r}
#explore data
head(medicaid_expansion)
summary(medicaid_expansion)

# highest and lowest uninsured rates prior to 2014

#create new variable for uninsured Americans per state
medicaid_expansion$pop_uninsured <- medicaid_expansion$uninsured_rate * medicaid_expansion$population

#bar plots for 2008-2013
#2008
bar_2008 <- ggplot(medicaid_expansion %>% filter(year == '2008'),
                   aes(State, uninsured_rate)) + geom_col() +
  ggtitle('Uninsured Rate Per State in 2008') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2008

#2009

bar_2009 <- ggplot(medicaid_expansion %>% filter(year == '2009'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2009') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2009

#2010
bar_2010 <- ggplot(medicaid_expansion %>% filter(year == '2010'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2010') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2010

#2011
bar_2011 <- ggplot(medicaid_expansion %>% filter(year == '2011'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2011') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2011

#2012
bar_2012 <- ggplot(medicaid_expansion %>% filter(year == '2012'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2012') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2012

#2013
bar_2013 <- ggplot(medicaid_expansion %>% filter(year == '2013'),
                   aes(State, uninsured_rate)) +
  geom_col() +
  ggtitle('Uninsured Rate Per State in 2013') +
  xlab('State') +
  ylab('Uninsured rate') +
  theme(axis.text.x = element_text(
    size = 8, angle = 90))
bar_2013


```
Prior to 2014, Massachusetts consisently had that lowest uninsured rate of all 
50 states (2008-2013), which is to be expected given that Massachusetts passed
a universal healthcare law in 2006. From 2008-2009, Utah had the highest 
uninsured rate, but that shifted in 2010-2013, with Texas having the highest 
uninsured rate. 

```{r}
# most uninsured Americans

minmax <- medicaid_expansion %>% group_by(year) %>% 
  summarize(min = min(pop_uninsured), 
            minState = State[which.min(pop_uninsured)],
            max =max(pop_uninsured), 
            maxState = State[which.max(pop_uninsured)])
View(minmax)
```

California, the most populous state, had the most uninsured Americans from 
2008-2013 (which makes sense given that California has the largest population
among all 50 states). After 2014 and the Medicaid Expansion, Texas replaced 
California as the state with the most uninsured Americans. Wyoming had the least
uninsured Americans from 2008-2009, but was replaced by Vermont from 2010
through 2010. 

# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.
- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot
#tried Colorado-Utah, Arizona-Utah, Idaho-Washington, Pennsylvania-New York,
# Virginia-Kentucky, and settled on Idaho-Nevada

medicaid_expansion %>%
  filter(State %in% c("Idaho","Nevada")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = pop_uninsured, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = pop_uninsured, 
                color = State)) +
  geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Uninsured population from 2008 to 2020') +
  xlab('Year') +
  ylab('Population Uninsured')
```
The parallel trends plot for the population uninsured over time between Idaho
and Nevada seems to meet the assumption that the trends between the two states
were fairly parallel prior to the treatment year, showing roughly similar slopes
between 2009 to 2013. 

- Estimate a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Difference-in-Differences estimation, uninsured population

# Nevada - Idaho
data_dd <- medicaid_expansion %>%
  filter(State %in% c("Nevada","Idaho")) %>%
  filter(year >= 2013 & year<= 2015) 

# pre-treatment difference, unuinsured pop

pre_diff_pop <- data_dd %>%
  filter(year == 2013) %>%
  select(State, 
         pop_uninsured) %>%
  spread(State, 
         pop_uninsured) %>%
  summarise(Nevada - Idaho)
  
# post-treatment difference, uninsured pop

post_diff_pop <- data_dd %>%
  filter(year == 2015) %>%
  select(State, 
         pop_uninsured) %>%
  spread(State, 
         pop_uninsured) %>%
  summarise(Nevada - Idaho)

# diff-in-diffs, uninsured pop

diff_in_diffs_pop <- post_diff_pop - pre_diff_pop
diff_in_diffs_pop


#Diff-in-diff, uninsured rate

#pre-treatment difference, unuinsured rate
pre_diff_rate <- data_dd %>%
  filter(year == 2013) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(Nevada - Idaho)
  
# post-treatment difference, uninsured rate

post_diff_rate <- data_dd %>%
  filter(year == 2015) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(Nevada - Idaho)

# diff-in-diffs, uninsured rate

diff_in_diffs_rate <- post_diff_rate - pre_diff_rate
diff_in_diffs_rate

```
Based on the differences-in-differences estimation for the uninsured population, 
the expansion of Medicaid had a treatment effect of -154374.4, or decreased the 
population of uninsured Americans by 154,374 people. Looking at the difference-
in-differences estimation for the uninsured rate, Medicaid expansion had a
treatment effect of -0.03, or a 3% decrease in the rate of uninsured Americans.

## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**: That intiuition is harder to apply to this data given that the 
units of interest were states, which are more difficult to compare to each other
because of differences in demographics, politics, economics, etc. 

- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?
- **Answer**: The strengths of using the parallel trends assumption is that it
provides a quick way to visually ascertain that the trends over a long period 
of time between treatment and control groups prior to the treatment are
generally constant. The weaknesses are that there is no way to statistically 
test this assumption and that it ignores possible confounding differences
between treatment and control groups (e.g., assuming that the treatment and 
control groups are similar enough on all possible covariates of interest and 
therefore their outcome trends are comparable).


# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

```{r}
# non-augmented synthetic control

#remove other states that adopted expansion, add in treated indicator for Louisiana
syn_la_df <- medicaid_expansion %>% 
  filter(is.na(Date_Adopted) == 1 | State == "Louisiana") %>% 
  mutate(adoption_year = as.numeric(as.character(as.Date(Date_Adopted, format = "%Y/%m/%d"), 
                                                 format = "%Y"))) %>% 
  mutate(treated = ifelse(State == "Louisiana" & adoption_year < year, 1, 0))

syn_louisiana <- augsynth(uninsured_rate ~ treated, State, year, syn_la_df,
                progfunc = "None", scm = T)

summary(syn_louisiana)

plot(syn_louisiana)
```
The average ATT estimate is -0.029. The L2 Imbalance is 0.008.

```{r}
syn_sum <- summary(syn_louisiana)

la_synla <- syn_la_df %>%
  filter(State == "Louisiana") %>%
  bind_cols(difference = syn_sum$att$Estimate) %>% #Estimate = Actual Uninsured Rate - Synthetic
  mutate(synthetic_louisiana = uninsured_rate - difference)

# Plot

la_synla %>%
  ggplot() +
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = 'Louisiana')) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = 'Louisiana')) +
  geom_point(aes(x = year, 
                 y = synthetic_louisiana, 
                 color = 'Synthetic Louisiana')) +
  geom_line(aes(x = year, 
                y = synthetic_louisiana, 
                color = 'Synthetic Louisiana')) +
  scale_color_manual(values = c('Louisiana' = 'red', 'Synthetic Louisiana' = 'blue')) +
  geom_vline(aes(xintercept = 2016)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Louisiana and Synthetic Louisiana') +
  xlab('Year') +
  ylab('Uninsured Rate')
```
The fit before 2016 looks pretty good, though we are a bit concerned that the pre-treatment fit starts coming apart right before the treatment. After 2016, we see that the uninsured rate for Louisiana was lower than it otherwise would have been without the Medicaid expansion.


- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
ridge_syn_la <- augsynth(uninsured_rate ~ treated, State, year, syn_la_df,
                progfunc = "ridge", scm = T)

summary(ridge_syn_la)
```
The Average ATT estimate is again -0.029. And the L2 Imbalane is again 0.008.

```{r}
ridge_syn_sum <- summary(ridge_syn_la)

ridge_la_synla <- syn_la_df %>%
  filter(State == "Louisiana") %>%
  bind_cols(difference = ridge_syn_sum$att$Estimate) %>% #Estimate = Actual Uninsured Rate - Synthetic
  mutate(synthetic_louisiana = uninsured_rate - difference)

# Plot

ridge_la_synla %>%
  ggplot() +
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = 'Louisiana')) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = 'Louisiana')) +
  geom_point(aes(x = year, 
                 y = synthetic_louisiana, 
                 color = 'Synthetic Louisiana')) +
  geom_line(aes(x = year, 
                y = synthetic_louisiana, 
                color = 'Synthetic Louisiana')) +
  scale_color_manual(values = c('Louisiana' = 'red', 'Synthetic Louisiana' = 'blue')) +
  geom_vline(aes(xintercept = 2016)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Louisiana and Synthetic Louisiana') +
  xlab('Year') +
  ylab('Uninsured Rate')
```

- Plot barplots to visualize the weights of the donors.

```{r}
# barplots of weights

##non-augmented synthetic control
# Convert weights to dataframe
data.frame(syn_louisiana$weights) %>%
  # change index to a column
  tibble::rownames_to_column('State') %>%
  ggplot() +
  # stat = identity to take the literal value instead of a count for geom_bar()
  geom_bar(aes(x = State, 
               y = syn_louisiana.weights),
           stat = 'identity') +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        axis.text.x = element_text(angle = 90)) +
  ggtitle('Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight') 


##ridge augmented synthetic control
# Convert weights to dataframe
data.frame(ridge_syn_la$weights) %>%
  # change index to a column
  tibble::rownames_to_column('State') %>%
  ggplot() +
  # stat = identity to take the literal value instead of a count for geom_bar()
  geom_bar(aes(x = State, 
               y = ridge_syn_la.weights),
           stat = 'identity') +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        axis.text.x = element_text(angle = 90)) +
  ggtitle('Augmented Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight') 

```

It looks like Florida was weighted quite heavily in the synthetic control, followed by South Carolina and South Dakota. 

**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states? Yes! We need to remove states that adopted the Medicaid expansion so that they will not be included as potential donor states to the synthetic control.

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?
- **Answer**: Instead of needing to find one state that can serve as an appropriate control group, we can rely on a set of states and (importantly) WEIGHT them based on their similarities to the state of interest along observable covariates. Put simply, it allows us to create a more appropriate control group. 

- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?
- **Answer**: Negative weights raise concerns about possibly unchecked extrapolation bias, because it uses extrapolation outside of the convex hull (N.B. I'm not actually sure that that means!) to generate the weights. Interpretation is also difficult, since it's not clear what it would mean for a state's data to negatively contribute to the construction of a synthetic control. To the extent we are not overly concerned about extrapolation outside the support of the data in traditional regression approaches and we are not that concerned about how different states were weighted in creating the synthetic control, we might be willing to allow for negative weights, especially if it noticeably enhances pre-treatment balance.

# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.


So we need a few thingfs for a Staggered Adoption Synthetic Control: 

- `year`, `State`: The state and year of the measurement
- `Date_Adopted`: The year that the state adopted medicaid expansion 
- `uninsured_rate`: rate of insured folks per state
- `treatsm`: whether or not the state got the treatment 


```{r}
medicaid_expclean <- medicaid_expansion %>%
   mutate(dateadopt =as.numeric(as.character(as.Date(Date_Adopted,format="%Y/%m/%d"),format = "%Y"))) %>%
   mutate(treatsm = 1 * (year >= dateadopt))

#was it treated that year onward 

#We're ready to start estimating a model! To do this, we use the #`multisynth()` function that has the following signature:
#mutltisynth(outcome ~ treatment, unit, time, nu, data,  n_leads)

# with a choice of nu
mpool_msyn <- multisynth(uninsured_rate ~ treatsm, State, year,
                        nu = 0.5, medicaid_expclean, n_leads = 10)
# with default nu
mpool_msyn <- multisynth(uninsured_rate ~ treatsm, State, year,
                        medicaid_expclean, n_leads = 10)

print(mpool_msyn$nu)

mpool_msyn

```

```{r}
mpool_msyn_summ <- summary(mpool_msyn)
```



```{r}
mpool_msyn_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = "bottom") +
  ggtitle('Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Uninsured Rate')
```

```{r}
mpool_msyn_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = 'None') +
  ggtitle('Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Uninsured Rate') +
  facet_wrap(~Level)
```

- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted epxansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# multisynth model time cohorts
mpool_msyn_tc <- multisynth(uninsured_rate ~ treatsm, State, year,
                            medicaid_expclean, n_leads = 10, time_cohort = TRUE)

mpool_msyn_tc_summ <- summary(mpool_msyn_tc)

mpool_msyn_tc_summ

```
```{r}
mpool_msyn_tc_summ$att %>%
  ggplot(aes(x = time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = "bottom") +
  ggtitle('Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Uninsured Rate')
```

#Now let's plot the Time Cohorts

```{r}
mpool_msyn_tc_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = 'None') +
  ggtitle('Multi-Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Uninsured Rate Per State') +
  facet_wrap(~Level)
```

## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?
- **Answer**: 
Yes, if we look at the example with the time cohorts model, for example, we can see that states that adopted medicaid expansion in 2014 vs. those in 2019 had different effect sizes. The effect of medicaid expansion on these states was quite different looking at the reducation in the rate of uninsured people. 


- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: 


There does some to be some evidence, albeit limited, that those who adopted early enjoyed a larger decreased in the uninsured population. For example, looking at the time cohorts, we can see that earlier adopters of medicaid expansion had an uninsured rate say those who adopted in 2015 or 2016 enjoyed decreases in uninsured rate enough to make their rates of uninsured folks be negative. I am not sure how to  interpret the negative rate in the context of the uninsured population but in general it seems like states that adotped much later had positive rates of uninsured and the trend says relatively consistent until adoption. 


# General Discussion Questions

- Why are DiD and  synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**:

First and foremost--many questions in social science are centered around aggregated units (e.g., cities, states) and thus, the panel data required to do DiD is readily available. In addition to data being readily available, it is also the case that many policy interventions concomitantly occur at the aggregate unit level. 

Synthetic control approaches are data-driven and thus, reduce the amount of discretion researchers have in making subjective decisions about control and treatment groups.Thus, synthetic control approaches make decisions around treatment and control groups more transparent using quantifiable (i.e. observable) characteristics to make treatment and control groups more comparable. 

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**:

There are two main issues in DiD/synthetic control: issue of selection across group and selection across time. For example, groups might change in composition over time, making the observed difference less valid in terms of a true calculation of differences between treatment and control groups. DiD/Synthetic control relies on the fact that the groups under study do not change systematically. The other issue of selection bias stems from unobservable characteristics that affect selection into treatment (e.g., entering into a program vs. not). 


What is advantageous about Regression Discontinuity in terms of selection bias is that the outcome of interest is observed given a specified exogenous cutoff, thus it follows logically that those above and beyond the cutoff are not dissimilar systematically. It is quite unlikely, for example, that those above and beyond an SAT threshold for a scholarship are systematically different from each other. The main issue that RD suffers from is external validity (i.e., it's hard to generalize to units not close to the cutpoint) while being close to the ideal conditions of an experiment. Thus, it's hard to generalize from an RD design but it's advantageous in terms of internal validity (i.e., identification). 

RD is useful when comparing groups that differ only in treatment (or when there is scientific and good rationale for assuming so). DiD is about comparing groups who not only differ in treatment but differ in other ways with the assumption that difference in outcomes is constant under the period of study. 